# Creating Analytic Datasets

In this chapter, we will work on manipulating and combining data to create analytic datasets.

## RQ1: What is the distribution of warm summer days in 2025 by North Carolina county?

In this research question, our goal is to create an analytic variable representing the percent of daily average summer temperatures that were above 70 degrees farenheit by NC county

### Data:

-   North Carolina county boundaries from US Census Bureau
-   2025 summer season daily average temperature values for each weather station in NC air quality data by station from [North Carolina State Climate Office](%22https://products.climate.ncsu.edu/cardinal/%22)

```{r, eval = F}
library(tidyverse)
library(sf)

nc_counties <- st_read("https://drive.google.com/uc?export=download&id=1g9sGIikgOEubqoj97fUVoCYAKlBDVX5a")

temp <- read_csv("https://drive.google.com/uc?export=download&id=16wJAatPKM0cF7VNy8hwNWOeJiiDNvsTt")


```

### Processing Steps:

-   In the `temp` object, each row represents a single day in the 2025 summer season (June - August) at a single weather station in North Carolina. Aggregate the `temp` object to the station level (this will summarize values at that station over the season). The aggregated dataset should include two calculated columns `total_obs` (total number of observations per station) and `total_over_70` (total observations per station where the daily temperature value is over 70 degrees). The `group_by` command must include `site`, `latitude`, and `longitude`
-   Spatialize the station-level data using the `st_as_sf()` command. The CRS = 4326.
-   Reproject the `nc_counties` and `temp` objects into CRS = 2264 (North Carolina State Plane, ft) (i.e. `nc_counties <- nc_counties |> st_transform(crs = 2264)`)
-   Execute a spatial join between `temp` and `nc_counties`. Drop geometry and aggregate the spatial `temp` object to the county level (take the sum of the `total_obs` and `total_over_70` columns per county).
-   Calculate a `pct_over_70` variable representing the percent of county observations that are over 70 degrees.
-   Use a table join to add your county-aggregated data to the `nc_counties` object.

## RQ2: How does tree cover canopy vary within walking distance from bus stops in Chapel Hill, NC?

In this research question, our goal is to create an analytic variable representing the percent tree canopy cover within .5 miles of each bus stop in Chapel Hill, NC.

### Data:

-   Bus stop locations from [Chapel Hill Open Data](https://opendata-townofchapelhill.hub.arcgis.com/datasets/townofchapelhill::chapel-hill-transit-bus-stops/explore)
-   Tree canopy cover from [NLCD](https://www.mrlc.gov/data/type/nlcd-tree-canopy-cover). Each pixel value represents the percent of tree canopy cover in that pixel.

```{r, eval = F}
library(terra)
library(exactextractr)

bus_stops <- st_read("https://drive.google.com/uc?export=download&id=1jRINUl-5uBAsBcWnKTRmdUO7ZTs1EV4G")

tree_canopy <- rast("https://drive.google.com/uc?export=download&id=1_SqO-ocyLCg3g1Mv1ZbqOd5Pa1sTddps")

```

### Processing Steps:

-   Transform `bus_stops` object into EPSG:3857 to match the tree canopy projection. Note that the units in this projection are meters
-   Buffer the `bus_stops` 804.672 meters (.5 miles)
-   Use the `exact_extract()` function to add a field to the buffered bus stop object that represents the average canopy cover
-   Create a simplified dataset that includes only the following fields: `STOP_ID`, average tree canopy variable

## RQ3: How accessible are bus stops to Chapel Hill addresses?

In this research question, our goal is to create an analytic variable representing the distance of each address in Chapel Hill to the nearest bus stop.

### Data:

-   Bus stop locations from [Chapel Hill Open Data](https://opendata-townofchapelhill.hub.arcgis.com/datasets/townofchapelhill::chapel-hill-transit-bus-stops/explore)
-   Chapel Hill addresses from [Chapel Hill Open Data](https://opendata-townofchapelhill.hub.arcgis.com/datasets/townofchapelhill::chapel-hill-transit-bus-stops/explore)

```{r, eval = F}
ch_addresses <- st_read("https://drive.google.com/uc?export=download&id=1fFXfEbOWjwfsT_JLeLYnaVPkbJCGYA2_")

```

### Processing Steps:

-   Transform the `ch_addresses` object into EPSG:3857 to match the bus stop object.
-   Calculate the distance (in meters) from each address to the nearest bus stop.
-   Create a simplified dataset that includes only the following fields: `OBJECTID`, `LBCSDesc`, distance variable
