# Basic Tidyverse Manipulation

This chapter will introduce you to the basics of working with the tidyverse. After completing these exercises, you should be able to:

-   Use the `mutate`, `filter`, `rename`, `summarise`, and `select` commands from the tidyverse package
-   Create a descriptive statistics table using the gt package
-   Create basic visualizations and describe data distributions (histogram, boxplot/violin plot, scatterplot, bar chart)

We will continue to use the data that we used in the Basic R chapter. To start, create a new .Rmd document (formatted like we formatted our .Rmd in the Basic R chapter). Read in the data using the same command we used in our previous .Rmd.

## Data Structure in the Tidyverse

Add a first-level header called "Using the Tidyverse". Under the header, add a code chunk

For the most part, we will be working with a special data structure called a `tibble`. A tibble is tidyverse's version of a data frame. In a tibble, each row is an observation and each column is a variable. This is called "tidy" data.

To get a basic understanding of our data, we can use the `str` command:

```{r, eval=FALSE}
#structure of absences
str(absences)

#structure of rurality
str(rurality)
```

You should also double-click each object in the "Environment" tab. This will open the tibble so that you can explore it.

**Q1: How many rows are in the absences dataset?**

**Q2: How many variables are in our rurality dataset?**

**Q3: What is the data type of the `year` variable in the absences dataset?**

**Q4: Which of the absences variables are characters?**

Basic data manipulation in the tidyverse generally involves "piping". Piping is the way that we can "chain" operations or commands and it looks like this: `|>` (in documentation you might also see a pipe written like this `%>%`). Run each of the following commands and add a descriptive comment above it.

```{r, eval=F}
#
absences |> slice_head(n = 5)

#
absences |> select(1)

#
absences |> select(rural)

#
absences |>
  summarise(mean_percent = mean(pre_covid_pct))

#
absences |>
  summarise(max_percent = max(after_covid_pct))

#
rurality |> summarise(n_rural = sum(nc_rural_center == "Rural"))

#
summary <- absences |>
  summarise(max_before_covid = max(pre_covid_pct), max_after_covid = max(after_covid_pct))
```

**Q5: Why wasn't the last command printed to the console?**

**Q6: Write a command that creates an object called `metro_core` and counts the number of observations that have a `ruca_code` of "Metropolitan Core"**

Usually, unless we are doing basic data exploration, we want to save our manipulated data as objects. You can either "re-write" an object by using the same name, or you can re-name the manipulated object. Run each of the following commands and add a descriptive comment above it.

```{r, eval=FALSE}
#
non_rural_data <- absences |> filter(rural == "non-rural")

#
pre_covid <- absences |> select(agency_code, pre_covid_pct)

#
renamed_absences <- absences |> rename(pre_covid_ratio = pre_covid_pct, after_covid_ratio = after_covid_pct)

#
added_column <- renamed_absences |> mutate(change = after_covid_ratio - pre_covid_ratio)

#
logical <- renamed_absences |> mutate(high_absence = ifelse(pre_covid_ratio > .25, 1, 0))

#
logical_rural <- rurality |> mutate(highly_rural = ifelse(ruca_code %in% c("Small town low commuting", "Rural area", "Small town core"), "Highly Rural", "Not Highly Rural"))
```

**Q7: Write a command that creates an object called `high_absence_schools` that filters the `logical`** **object to just schools that have a high absence rate before Covid.**

**Q8: Write a command that creates an object called** `urban` **that filters the `rurality` object to just observations that are "Metropolitan high commuting" or "Metropolitan core"**

## Descriptive Statistics

Descriptive statistics allow us to understand the distribution of our dataset. We can use a similar command structure to create almost all of the descriptive statistics tables in our class. Add a first-level header called "Describing the Data". Add a code chunk below.

```{r, eval = F}
##Notice that you need to add these libraries. You might need to install them. 
library(e1071)
library(gt)

#calculate descriptive statistics 
pre_covid_stat <- absences |> select(pre_covid_pct) |>
  summarise(
    n = n(),
    num_na = sum(is.na(pre_covid_pct)), 
    mean_pct = mean(pre_covid_pct, na.rm = TRUE),
    median_pct = median(pre_covid_pct, na.rm = TRUE),
    sd_pct = sd(pre_covid_pct, na.rm = TRUE),
    variance_pct = var(pre_covid_pct, na.rm = TRUE),
    mean_dev = mean(abs(pre_covid_pct - mean(pre_covid_pct, na.rm = TRUE)), na.rm = TRUE),
    cv_pct = sd_pct / mean_pct,
    min_pct = min(pre_covid_pct, na.rm = TRUE),
    max_pct = max(pre_covid_pct, na.rm = TRUE),
    skewness = skewness(pre_covid_pct, na.rm = TRUE), 
    kurtosis = kurtosis(pre_covid_pct, na.rm = TRUE)
  )

##this will format the table for display
display_table <-  pre_covid_stat |> pivot_longer(everything(), names_to = "Statistic", values_to = "Value") |>
  gt() |>
  tab_header(
    title = "Chronic Absences at NC Public Schools before Covid-19",
  ) %>%
  fmt_number(
    columns = everything(),
    decimals = 2
  )

#display the table
display_table


### CATEGORICAL DATA #####
## create a basic table of descriptive statistics for NC Rural Center Rural Definition
rural_table <- rurality |> 
  group_by(nc_rural_center) |>
  summarise(n = n())  |>  gt() 

rural_table
```

## Basic Plotting with ggplot2

Data visualizations can add additional information to our descriptive statistics table. Add a first-level header called "Plotting our Data". Add a code chunk below it.

### The "Language" of ggplot2

The ggplot2 package builds visualizations using a conceptual framework based on the grammar of graphics– meaning that every graph can be built from the same components. These components are a dataset, a coordinate system, and geoms (visual marks that represent data points). The basic "formula" of a ggplot is:

1\) Start with the data -\> ggplot(DATAFRAME,

2\) Map variables to aesthetics -\> aes(x = , y =, color = , size = , shape= )

3\) Add layers (what kind of graph) -\> + geom\_\*\*\*

For instance, a basic scatterplot could look like

`ggplot(data, aes(x = var_for_x_axis, y = var_for_y_axis)) + geom_point()`

We will work extensively with ggplot2 graphs in this class. The following examples provide a very brief introduction on using ggplot2

### Histogram/ Density Plot

A histogram is a graph that displays the frequency of observations within user-set "bins". A density plot is a smooth curve that shows the distribution of the data. The curve represents the proportion of the data in each range rather than the frequency. Both of these visualizations help us understand the distribution of our data

```{r, eval = F}
## create a histogram of pre covid absences
#absences is our data, x is the variable for x axis (no y axis with histograms, since it is frequency)
#binwidth is how large each bin is
#alpha sets the transparency
#labs adds labels
ggplot(absences, aes(x = pre_covid_pct)) + 
  geom_histogram(binwidth = .05, fill = "lightgrey", color = "black", alpha = 0.7) +
  labs(
    title = "Histogram of Pre Covid Absences",
    x = "Absence Ratio",
    y = "Frequency"
  ) 

## create a density plot of pre covid absences
ggplot(absences, aes(x = pre_covid_pct)) +
  geom_density(fill = "lightgrey", color = "black", alpha = 0.7) +
  labs(
    title = "Density Plot of Pre Covid Absences",
    x = "Absence Ratio",
    y = "Proportion"
  ) 
```

#### Boxplot

A boxplot is a standardized way of displaying the distribution of data based on a five-number summary: minimum, first quartile (Q1), median, third quartile (Q3), and maximum. It can also highlight outliers in the data.

```{r, eval = F}
## absences is the data
## y axis only (if you wanted to flip the box plot, you would use x = )
ggplot(absences, aes(y = pre_covid_pct)) +
  geom_boxplot(fill = "lightgrey") +
  labs(
    title = "Boxplot of Chronic Absences",
    y = "Ratio"
  )
```

**Q16: Make a box plot of after covid absences**

#### Violin Plot

A violin plot is a method of plotting numeric data and can be understood as a combination of a boxplot and a density plot. It shows the distribution of the data across different values.

```{r, eval = F}
##Look how we've layered two different plot types onto the same graph
ggplot(absences, aes(x = 1, y = pre_covid_pct)) +
  geom_violin(fill = "lightgrey", bw = .05)  + geom_boxplot(width=0.3) +
  labs(
    title = "Violin Plot of of Chronic Absences",
    y = "Ratio"
  ) + theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

### Bar Chart

A bar chart displays the count of each observations in a group. It is a useful visualization for nominal/categorical data.

```{r, eval = F}
#this creates the grouped dataset
grouped_rural <- rurality |> 
  group_by(ruca_code) |>
  summarise(Count = n()) 

#this plots the bar plot (note that the coord_flip swtiches the axis which is better for labeling in this case)
ggplot(grouped_rural, aes(x=ruca_code, y=Count)) + 
  geom_bar(stat = "identity") + coord_flip() + labs(
    x = "RUCA Code Category",
    y = "Number of Census Tracts",
  )
```

### Faceting and Grouping Plots

Faceting and grouped visualization are two ways to compare distributions across categories in `ggplot2`. **Faceting** creates separate panels for each level of a categorical variable, making it easy to see each group individually. **Grouped visualization** overlays multiple groups in a single plot, using color or fill to distinguish them, so you can compare distributions directly. In both cases, the variable used to define the groups must be **categorical** (like `rural`), because these techniques rely on discrete levels to separate or color the data.

```{r, eval = F}
##grouped visualization (see fill value)
ggplot(absences, aes(x = pre_covid_pct, fill = rural)) +
  geom_density(binwidth = .05, alpha = .7) +
  labs(
    title = "Histogram of Pre Covid Absences",
    x = "Absence Ratio",
    y = "Frequency"
  ) 

#faceted visualization (see facet_wrap)
ggplot(absences, aes(x = pre_covid_pct)) +
  geom_density(binwidth = 0.05, alpha = 0.7, fill = "lightblue") +
  facet_wrap(~ rural) +   # create separate panels by the 'rural' variable
  labs(
    title = "Density of Pre-Covid Absences by Rural Category",
    x = "Absence Ratio",
    y = "Density"
  )

```

```{r, eval = F}
#violin plots and box plots use the x variable to look at categories
ggplot(absences, aes(x = rural, y = pre_covid_pct)) +
  geom_violin(fill = "lightgrey", bw = .05)  + geom_boxplot(width=0.1) +
  labs(
    title = "Violin Plot of Summer Temperatures (1991-2020)",
    y = "Summer Temperature (°F)"
  )  


```

## Mini-Challenge

Add a new first level header called "Mini-Challenge #2". Under that header, add a code chunk. In the code chunk, write commands to do the following:

-   Make a descriptive statistics table of `after_covid_ratio`
-   Make a descriptive statistics table of `ruca_code`
-   Make a histogram and density plot of after covid absences
-   Make a violin chart of after covid absences
-   Make a bar chart of NC Rural Center rural definition
